---
title: "Recitation 10"
author: "Eugene Tan (TA)"
date: "11/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## This class

* Some comments on HW2
  * how to google
  * data.table
* Date handling
  * lubridate
* Fixed Effects models
  * plm
  * lfe

## HW2 - how to google

Always google with "data.table R thing_you_want_to_google"

* Tidyverse vs data.table
  * because most R users use tidyverse, googling will normally by default give you tidyverse results
  * we've been teaching data.table in this class
  * NOT COMPATIBLE - running a tidyverse operation on a data.table means you have to use as.data.table again
  * Multiple versions because people wanted different ways of doing abstract things
* Pdata (from plm) is another type of dataframe like tibbles (tidyverse), data.table.

## Data.table
dt[i,j, by]

* i filters or selects a subset of data (by rows) that you want
* j selects or does operations on multiple columns
* by does things on different categories of the data. 

## Lubridate converts strings to POSIXct
```{r warning = FALSE}
library(lubridate)
# Change STRINGS to dates
dmy("20-08-2008")
mdy("08/20/2008")
ymd("200820")
```

## Lubridate converts strings to POSIXct
```{r}
ymd("190820")
ymd("20.12.11")
ymd("2020 November 4th")
mdy('Nov 4 2020')
mdy('Nov042020')
```

## Convert things into string
```{r echo = TRUE}
# paste converts numerics to strings
paste(1994, "10", 1, sep = "-0")
paste(2022, "10", 1)
paste0(1994, "10", 1)
paste0(2022, "10", 1)
?paste
```

## Combining them together
```{r}
y <- 2020
d <- "30st"
m <- 4
# Pseudocode
# Combine y, m, and d into a string
date_string <- paste(y,d,m, sep = '-')
# Convert string to date
ydm(date_string)
```


## Combining them together
```{r warning = TRUE}
y <- 2020
d <- "1st"
m <- 11
# Pseudocode
# Combine y, m, and d into a string
ex <- paste(y,d,m)
# Convert string to date
ydm(paste(y,d,m))

?strptime
```


```{r}
# One-liner
ydm(paste(y,d,m, sep = "-"))
ydm(paste0(y,m,d)) # This fails why?
```

## Parsing dates on data.table (setup)
Remember that data.table is vectorized
```{r warning  = FALSE}
library(data.table)
?airquality
aq <- as.data.table(airquality)
year <- 1973
head(aq)
```
## Doing this on data.table
Remember that data.table is vectorized
```{r}
# dt[i,j,by]
# Use paste in vectorized form
head(aq[,paste(year,Month, Day)])
# Converting to date
head(aq[,ymd(paste(1973, Month, Day))])
# Assigning to variable
aq[, date := ymd(paste(1973, Month, Day))]

# First measurement in August where the Solar Radiation is larger than 240?
# Filter August 
# Filter for the solar radiation larger than 240
# Find the first row
setorder(aq, date)
aq[(Month==8)&(Solar.R > 240)]
aq[Month==8][Solar.R > 240]
```

## Doing it outside data.table
```{r}
# Lets get the vector only
d <- aq[, Day]
m <- aq$Month
# Show that we only have the vector
head(d)
# Use the paste with dmy
head(dmy(paste(d,m,1973)))
# Show that we get a vector back
class(dmy(paste(d,m,1973)))
# Put the vector back in
aq$date2 <- dmy(paste(d,m,1973))
aq[, date2 := dmy(paste(d,m,1973))]
```

## Load the PS3 dataset
```{r}
# Removes all variables from my environment (helpful to put this at the top of your script)
rm(list=ls()) 
# Load my version of the clean data.
dt <- readRDS("../homeworks/hw3/data_clean.Rds")
# (Save using saveRDS(data, "data_clean.Rds"))
dt
```

## PLM vs LFE

These are alternatives or substitutes to deal with panel data - both provide support for:

* fixed effect regressions
* clustered standard errors
* diff-in-diff regressions
* regression discontinuity

PLM pluses

* More flexible (but more configuration) handling of dates and groups

LFE pluses

* Better handling for weights (I'll show the weighted regression)
* You might want the degress of freedom correction when you run  [clustered standard errors](https://stackoverflow.com/questions/30116099/clustered-standard-errors-different-in-plm-vs-lfe)

## plm Setup

```{r warning = FALSE}
library(plm)
library(stargazer)
# Look at the reactorid and ym
head(dt[,.(reactor_id, yearmonth)])
```

## plm 
```{r}
# Set those as the index/key
pdata <- pdata.frame(dt, c("reactor_id", "yearmonth"))

# Reactor fixed effects, age cubed. (Within == fixed effects)
r1 <- plm(capacity_factor ~ divested + as.factor(yearmonth) + age3, model = "pooling", pdata)
r2 <- plm(capacity_factor ~ divested + as.factor(yearmonth), model = "within", pdata)
#coeftest(r1, vcov=vcovHC(r1, cluster = "group"))

# Omit the fixed effects in stargazer and fix the standard errors for plm
stargazer(r1, r2, 
          type = 'text', 
          omit = c("yearmonth"), 
          se =  list(sqrt(diag(vcovHC(r1,  cluster = "group"))), 
                     sqrt(diag(vcovHC(r2,  cluster = "group")))),
          omit.stat=c("adj.rsq","f"))
```

## LFE Example
```{r warning = FALSE}
library(lfe)
# How to use the felm function:
# first part: formula without fixed effects
# second part: fixed effects
# third part: IV specification (write 0 if not using an IV)
# fourth part: level at which to cluster the standard errors
# This function allows to clustered standard errors at a level different from the fixed effects.
r3 <- felm(capacity_factor ~ divested + age3 | as.factor(yearmonth) + reactor_id  | 0 | plant_id, dt)
r4 <- felm(capacity_factor ~ divested + age3 | as.factor(yearmonth) + reactor_id  | 0 | plant_id, dt, weights = dt$capacity)

stargazer(r1,r2,r3,r4, 
          type = 'text',  
          omit = c("yearmonth", "age", "Constant"),
          se =  list(sqrt(diag(vcovHC(r1,  cluster = "group"))), 
                     sqrt(diag(vcovHC(r2,  cluster = "group")))),
          add.lines = list(c("R package", "plm","plm","lfe","lfe"),
                           c("Month-year FE","Yes", "Yes","Yes", "Yes"),
                           c("ID Fixed Effects","No","Reactor","Reactor","Reactor"),
                           c("Capacity Weights", "No","No","No", "Yes"),
                           c("Reactor age (cubic)", "Yes", "No", "Yes","Yes"),
                           c("#Cross sectional units", "103","103","103","103")),
          omit.stat=c("adj.rsq","rsq","f", "ser"))
```

## Hint for the plant level
Using PLM you'll want to change the index from reactor id to plant id. 


## Help 

```{r}

dt[, .G, ]
dt[, .N, manufacturer]
```

