---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

# Regression Example

## mtcars

This is a quick and dirty example of how to do regressions in r and use stargazer. 
I'm going to use the mtcars dataset, which comes [prepackaged with R](http://www.sthda.com/english/wiki/r-built-in-data-sets). From the documentation `?mtcars`, you can see the variable descriptions:

> A data frame with 32 observations on 11 (numeric) variables.  
[, 1]	mpg	Miles/(US) gallon  
[, 2]	cyl	Number of cylinders  
[, 3]	disp	Displacement (cu.in.)  
[, 4]	hp	Gross horsepower  
[, 5]	drat	Rear axle ratio  
[, 6]	wt	Weight (1000 lbs)  
[, 7]	qsec	1/4 mile time  
[, 8]	vs	Engine (0 = V-shaped, 1 = straight)  
[, 9]	am	Transmission (0 = automatic, 1 = manual)  
[,10]	gear	Number of forward gears  
[,11]	carb	Number of carburetors  

```{r, echo = TRUE}
?mtcars
library(data.table)
mtcars
dt <- as.data.table(mtcars)
```


## Formula syntax

First we have to talk about formulas. 

Formulas are a little bit of a new syntax, and you can get a pretty in depth tutorial from datacamp [here if you are interested](https://www.datacamp.com/community/tutorials/r-formula-tutorial).

But for empirical exercise 1 it'll be very simple - you only have two variables (carbon emissions and coal capacity)! The syntax for this regression (as above in the ) 

$$y = \alpha + \beta x + \epsilon$$
is just `y~x`. Let's see how we use it in base r.

## lm for linear model

The simplest regressions is linear regression, and on the mtcars dataset we'll run a regression of the miles per gallon on weight. 

The regression I run is this:

$$mpg = \alpha +\beta wt + \epsilon$$
and it's just this code.  The first parameter to the function lm is a *formula*, which is the syntax I describe above, and the second is the dataset. You can of course look at the other parameters using `?lm`. 

```{r}
lm(mpg ~ wt, dt)

```
As you can see, the $\beta_0$ term is 37, and the $\beta_1$ term is -5.  We interpret this is as -5.344 miles per gallon per 1000 lbs, which I got from the data descriptions.

If I wanted to add more variables, say a dummy for the transmission and the number of cylinders:

$$mpg = \alpha + \beta_0wt + \beta_1cyl + \beta_2am + \epsilon$$
Then my code would be `mpg ~ wt + cyl + am`. Let's try it:

```{r}
lm(mpg ~ wt + cyl + am, dt)
```

You might want to get more than just your coefficients, like standard errors. You can report more model output by saving the output to a variable and then using the function `summary()`. This way, we can see that in the first regression, weight is a significant predictor of mpg at the 99.9% level. In the second regression weight and cylinders are significant predictors at the 99% level, but transmission is not significant. 

```{r}
model1 <- lm(mpg ~ wt, dt)
summary(model1)
model2 <- lm(mpg ~ wt + cyl + am, dt)
summary(model2)
```


### Robust Standard Errors and panel data

Unfortunately, base r does not give robust standard errors by default. We use robust standard errors to adjust for [heteroskedasticity](https://en.wikipedia.org/wiki/Heteroscedasticity). 

First, install estimatr using `install.packages("estimatr")`. Now load the library, you'll use the [lm_robust](https://declaredesign.org/r/estimatr/) function which uses the same format as lm above, but with heteroskedasticity-consistent standard errors. (HC2 stands for heteroskedasticity consistent method 2).


```{r}
# The first model
library(estimatr)
model1 <- lm_robust(mpg ~ wt, dt)
model1
summary(model1)
```
Note we've lost our stars!

```{r}
# The second model
model2 <- lm_robust(mpg ~ wt + cyl + am, dt)
model2
summary(model2)
```

## Stargazer

Stargazer doesn't come prepackaged with R, so you need to install it first as well.
Run `{r} install.packages("stargazer")` to install it. 

Unfortunately, stargazer and estimatr aren't perfectly compatible, and so stargazer [has to be tricked](https://declaredesign.org/r/estimatr/articles/regression-tables.html#stargazer) into using robust standard errors.  

```{r}
library(stargazer)
# Make the fit the base r way
fit_1 <- lm(mpg ~ wt, data = mtcars)
fit_2 <- lm(mpg ~ wt + cyl + am, data = mtcars)
# use starprep (from the estimatr library) to uses lm_robust defaults for standard errors (robust HC2 SEs are the default in both lm_robust). 
# See https://declaredesign.org/r/estimatr/articles/regression-tables.html#stargazer

# Then compare the two regressions.
stargazer(fit_1, fit_2, se = starprep(fit_1, fit_2), type = 'text')
```





